# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data


```{r, echo=TRUE}

    setwd("C:/Users/Dan/Desktop/Coursera/5. Reproducible Research/Proj1") # set working directory

    if (!file.exists("activity.zip")) stop("Zip file missing") # check if file exists  
    activ_dat <- read.csv(unz("activity.zip", "activity.csv"),  header=T) # read in data
    activ_dat_no_na <- activ_dat[!is.na(activ_dat[,1]),] # removing obs. with missing no.of steps
  
```


## What is mean total number of steps taken per day?

```{r, echo=TRUE, fig.height=4}

    steps_day <- aggregate(steps ~ date, data=activ_dat_no_na, sum) # aggregate steps by day
    hist(steps_day$steps, breaks=10, main ="Daily Number of Steps (Histogram)", xlab="Number of Steps", ylab="Frequency of Daily Number of Steps") # create histogram

    steps_day_med <- round(median(steps_day$steps),0) # median rounded to integer value
    steps_day_mean <- round(mean(steps_day$steps),0) # mean rounded to integer value

```

- The average number of steps per day (October-November) reported for the given individual was 
`r steps_day_mean` while the median number of steps was `r steps_day_med`.

```{r, echo=TRUE, fig.height=4, results="hide"}

    ## Export plot to png

    dev.copy(png,'fig1_hist_days.png', width=800,height=600)
    dev.off()

```


## What is the average daily activity pattern?

```{r, echo=TRUE, fig.height=4}

    steps_interval <- aggregate(steps ~ interval, data=activ_dat_no_na, mean) # aggregate steps by interval
    plot(steps_interval$interval,steps_interval$steps,type="l", main = "Average Number of Steps per 5-min Interval", xlab = "5-min Interval", ylab = "Average Number of Steps")

    steps_interval_max <- steps_interval[steps_interval$steps==max(steps_interval$steps),1]

```

- On average, the 5-min time interval with the highest mean average number of steps was `r paste(steps_interval_max,"--",steps_interval_max+5)`.

```{r, echo=TRUE, fig.height=4, results="hide"}

    ## Export plot to png

    dev.copy(png,'fig2_tseries_interval.png', width=800,height=600)
    dev.off()

```

## Imputing missing values

```{r, echo=TRUE, fig.height=4}

    activ_dat2 <- read.csv(unz("activity.zip", "activity.csv"),  header=T) # re-read in data in new file 
    na_count <- sum(is.na(activ_dat2[,1]))  ## count obs. with missing steps
    na_index <- which(is.na(activ_dat2[,1])==TRUE) ## retain index of obs with missing steps
  
    for (i in 1:na_count) {
         activ_dat2[na_index[i],1] <- steps_interval[(steps_interval$interval==activ_dat2[na_index[i],3]),2]
         ## assign obs with missing steps the corresponding mean interval step values over the timeframe 
    }

    ##  Data with filled missing information

    steps_day2 <- aggregate(steps ~ date, data=activ_dat2, sum) # aggregate steps by day
    hist(steps_day2$steps, breaks=10, main ="Modified - Daily Number of Steps (Histogram)", xlab="Number of Steps", ylab="Frequency of Daily Number of Steps") # create histogram

    steps_day_med2 <- round(median(steps_day2$steps),0) # median rounded to integer value
    steps_day_mean2 <- round(mean(steps_day2$steps),0) # mean rounded to integer value

```

- The average number of steps per day (October-November) when missing data is filled in and that was reported for the given individual was `r steps_day_mean2` while the median number of steps was `r steps_day_med2`. The no-missing data mean was `r steps_day_mean` and the medin was `r steps_day_med`. Adding in mean values for the missing steps has no significant inpact on the daily distribution of steps other than increasing the frequency of the very centre of the distribution (due to adding days with totals that fall straight in the centre of the distribution). The mean does not change at all while the median changes ever so slightly, getting even closer to the mean.

```{r, echo=TRUE, fig.height=4, results="hide"}

    ## Export plot to png

    dev.copy(png,'fig3_hist_interval_nomiss.png', width=800,height=600)
    dev.off()

```

## Are there differences in activity patterns between weekdays and weekends?

```{r, echo=TRUE, fig.height=4}
    
   # Note: Original data is used; na's (steps) are removed   

   activ_dat[,4] <- weekdays(as.Date(activ_dat$date)) # create column of day names based on date column
   names(activ_dat)[4] <- "DayName"
   activ_dat_no_na <- activ_dat[!is.na(activ_dat[,1]),] # removing obs. with missing no.of steps
   
   activ_dat_no_na[activ_dat_no_na$DayName %in% c("Monday","Tuesday","Wednesday","Thursday","Friday"),5] <- "weekday"
   activ_dat_no_na[activ_dat_no_na$DayName %in% c("Saturday","Sunday"),5] <- "weekend" # split data by weekday, weekend
   names(activ_dat_no_na)[5] <- "TypeOfDay" # creata type of day variable and assign it a name 


   ## Aggrgate and plot

   totals <- aggregate(steps ~ interval + TypeOfDay, data=activ_dat_no_na, FUN = mean)

   library(ggplot2) ## initiate library

   ## initiate ggplot object and add in features

   gplot <- ggplot(totals, aes(interval, steps, group=TypeOfDay)) 
   gplot <- gplot + facet_grid(TypeOfDay ~ .) +geom_line(color="blue",size=1)
   gplot <- gplot + labs(title = "Average Number of Steps Taken Across Days")
   gplot <- gplot + xlab("Interval") + ylab("Average Number of Steps")

   print(gplot)

```

```{r, echo=TRUE, fig.height=4, results="hide"}


   ## Export plot to png

   dev.copy(png,'fig4_panel_days.png', width=800,height=600)
   dev.off()

```